\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-03-07 jasa/paper-jasa.tex
minor changes

As a simple and illustrative example, let us first consider the approximation of
the likelihood log-ratio $\log \left( r(\mathbf{x};\theta_0,\theta_1) \right)$ between the 1D mixtures
$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$ illustrated in
Figure~\ref{fig:1a} and defined as
\begin{align}
p(\mathbf{x}|\theta_0) &= (\frac{1}{2} - \gamma) p_{c_0}(\mathbf{x}) + (\frac{1}{2} - \gamma) p_{c_1}(\mathbf{x}) + \gamma p_{c_2}(\mathbf{x}), \\
p(\mathbf{x}|\theta_1) &= \frac{1}{2} p_{c_0}(\mathbf{x}) + \frac{1}{2} p_{c_1}(\mathbf{x}),
\end{align}
where $p_{c_0} := {\cal N}(\mu=-2, \sigma^2=0.5625)$, $p_{c_1} := {\cal N}(\mu=0, \sigma^2=4)$,
$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$ and where $\gamma$ is a parameter of $p(\mathbf{x}|\theta_0)$
set to $0.05$. Samples drawn from $p(\mathbf{x}|\theta_0)$ are shown in the
figure and will be used later for inference.

@@

As shown in the figures, qualitatively different results are obtained depending
the classification algorithm embedded within the approximation of the ratio. The
neural network-based approximation appears to work best, due to the capacity of
the classifier to model a smooth decision function $s(\mathbf{x})$ which is
very close to the optimal decision function. The random forest yields a decision function
which is piece-wise constant, therefore non-continuous and not strictly
monotonic with the ratio in the finite sample regime. On this problem, the
random forest still approximates the ratio correctly, but more variance can be
observed. Finally, the linear model produces a decision function that is
smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
the ratio around $\mathbf{x}=1$, this still results in an accurate
approximation, but poorer results should be expected on more complex problems.
While not directly apparent, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
s(\mathbf{x}))$ is key to obtain accurate results. Standard
histograms with fixed bins have been used here for illustrative purposes, but we
anticipate that the embedded density estimation algorithm should be carefully
chosen and tuned, on a case by case basis.

@@

Let us now consider the composite hypothesis testing setting, by comparing
$H_0: \gamma = 0$ versus the alternative $H_1: \gamma > 0$. That is, we want to evaluate
\begin{equation}
\Lambda({\cal D}; \Theta_0 = \{0\}, \Theta=(0, \frac{1}{2}]) =  \frac{ p({\cal D} | \gamma = 0) }{ \sup_{\theta \in \Theta} p({\cal D} | \gamma=\theta)} \; .
\end{equation}
Following the approach outlined in
Section~\ref{sec:generalized-likelihood-ratio}, computing the generalized
likelihood ratio statistic requires finding the maximum likelihood estimator
(MLE) for the denominator, which can be carried out using approximated
likelihood ratios. Reusing the decomposed neural network-based approximation and
minimizing the mean negative log-ratio of some observed data generated for
$\gamma=0.05$, Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,
compared to the distribution of the true MLEs. The figure clearly suggests that
the proposed inference procedure yields in this case an unbiased estimator
($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
trials). Note that had the underlying ratio not been accurate enough because of
poor training or poor calibration, the two distributions would not be exactly
overlapping, indicating estimator bias. As a second check, Figure \ref{fig:2b} also confirms
that the generalized likelihood ratio statistic obtained with our method is $\chi^2$ distributed,
as should be the true statistic.

>>>>

As a simple and illustrative example, let us first consider the approximation of
the likelihood log-ratio $\log \left( r(\mathbf{x};\theta_0,\theta_1) \right)$ between the 1D mixtures
$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$ illustrated in
Figure~\ref{fig:1a} and defined as
\begin{align}
p(\mathbf{x}|\theta_0) &= (\frac{1}{2} - \frac{\gamma}{2}) p_{c_0}(\mathbf{x}) + (\frac{1}{2} - \frac{\gamma}{2}) p_{c_1}(\mathbf{x}) + \gamma p_{c_2}(\mathbf{x}), \\
p(\mathbf{x}|\theta_1) &= \frac{1}{2} p_{c_0}(\mathbf{x}) + \frac{1}{2} p_{c_1}(\mathbf{x}),
\end{align}
where $p_{c_0} := {\cal N}(\mu=-2, \sigma^2=0.5625)$, $p_{c_1} := {\cal N}(\mu=0, \sigma^2=4)$,
$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$ and where $\gamma$ is a parameter of $p(\mathbf{x}|\theta_0)$
set to $0.05$. Samples drawn from $p(\mathbf{x}|\theta_0)$ are shown in the
figure and will be used later for inference.

@@

As shown in the figures, qualitatively different results are obtained depending
the classification algorithm embedded within the approximation of the ratio. The
neural network-based approximation appears to work best, due to the capacity of
the classifier to model a smooth decision function $s(\mathbf{x})$ which is
very close to the optimal decision function. The random forest yields a decision function
which is piece-wise constant, therefore non-continuous and not strictly
monotonic with the ratio in the finite sample regime. On this problem, the
random forest still approximates the ratio correctly, but more variance can be
observed. Finally, the linear model produces a decision function that is
smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
the ratio around $\mathbf{x}=1$, this still results in an accurate
approximation, but poorer results should be expected on more complex problems.
While not directly apparent, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
s(\mathbf{x}))$ is key to obtain accurate results. Standard
histograms with fixed bins have been used here for illustrative purposes, but we
anticipate that the embedded density estimation algorithm should be carefully
chosen and tuned, on a case by case basis. Overall, this first example confirms
that even if the classifier does a poor joob a modeling the optimal decision
function, good approximations of the likelihood ratio can still be obtained
with proposed method, by calibrating $s(\mathbf{x})$ and/or by decomposing
the mixture when possible.

@@

Let us now consider the composite hypothesis testing setting, by comparing
$H_0: \gamma = 0.05$ versus the alternative $H_1: \gamma \neq 0.05$. That is, we want to evaluate
\begin{equation}
\Lambda({\cal D}; \Theta_0 = \{0.05\}, \Theta=[0, 1]) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\theta \in \Theta} p({\cal D} | \gamma=\theta)} \; .
\end{equation}
Following the approach outlined in
Section~\ref{sec:generalized-likelihood-ratio}, computing the generalized
likelihood ratio statistic requires finding the maximum likelihood estimator
(MLE) for the denominator, which can be carried out using approximated
likelihood ratios. Reusing the decomposed neural network-based approximation from Figure~\ref{fig:1c} and
minimizing the mean negative log-ratio of observed data generated for
$\gamma=0.05$, Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,
compared to the distribution of the true MLEs. The figure clearly suggests that
the proposed inference procedure yields in this case an unbiased estimator
($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
trials). Note that had the underlying ratio not been accurate enough because of
poor training or poor calibration, the two distributions would not be exactly
overlapping, indicating estimator bias. As a second check, Figure \ref{fig:2b} also confirms
that the generalized likelihood ratio statistic obtained with our method is $\chi^2$ distributed,
as should be the true statistic when the null is true.
